To be FAIR: Theory Specification Needs an Update

### 2025-02-20 manuscript.Rmd
closes #51

Finally, metadata should be registered or indexed in a searchable registry (F4).
Zenodo and GitHub are both searchable.
Standardized metadata further improve Findability in these repositories.
The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
Furthermore, a standard keyword can be used; we suggest using the keyword `"fairtheory"` for all resources that constitute or reference a FAIR theory.
<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->

@@

One remaining issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":  <!--AB very important point that I was already missing in the introduction; seems we also need FAIR measurements :) --> <!-- NvD: I agree. Also, if we mention it here, we should definitely not mention it under limitations ->
it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].
Additionally, the "jingle-jangle" fallacy is prevalent in the social sciences:
the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
FAIR theory can help address the measurement crisis:
since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of

FAIR theory incorporates theory into open science workflows,
facilitates scholarly communication about theories,
making it easier to share theories with less opportunity for ambiguity and misunderstanding.
FAIR Theories are easier to find; they facilitate sharing, reusing, and updating open theories.
More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition,
removes barriers to knowledge exchange with the global scholarly community,
opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration.

>>>>

Finally, metadata should be registered or indexed in a searchable registry (F4).
It is important to note that, while may archives are technically searchable (e.g., GitHub, FigShare, the Open Science Framework, institutional repositories),
only few are specifically designed for FAIR-compliant archival.
Zenodo stands out in this respect, and it is in turn [indexed by Google Scholar](https://support.zenodo.org/help/en-gb/29-indexing/61-is-zenodo-indexed-by-google-scholar), Google Dataset Search, and other services.
Thus, while it is fine (and recommended) to use GitHub for version control,
the version of record should be archived in a FAIR repository like Zenodo.
Using standardized metadata further improves the Findability of theories archived within FAIR repositories.
The DataCite Metadata Schema provides a controlled vocabulary for research output, and the `resource_type: model` matches the description of FAIR theory [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- > An abstract, conceptual, graphical, mathematical or visualization model that represents empirical objects, phenomena, or physical processes. -->
Furthermore, we suggest using the keyword `"fairtheory"` for all resources that constitute or reference (a) FAIR theory.
<!-- (separating the words `FAIR` and `theory` by a space or hyphen would lead them to be interpreted as separate tokens in many search engines). -->

@@

One issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":
it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].
Additionally, the "jingle-jangle" fallacy is prevalent in the social sciences:
the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
FAIR theory can help address the measurement crisis:
since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of measurement.

<!-- FAIR theory incorporates theory into open science workflows, -->
<!-- facilitates scholarly communication about theories, -->
<!-- making it easier to share theories with less opportunity for ambiguity and misunderstanding. -->
<!-- FAIR Theories are easier to find; they facilitate sharing, reusing, and updating open theories. -->
<!-- More efficient and transparent communication about theory democratizes and accelerates cumulative knowledge acquisition, -->
<!-- removes barriers to knowledge exchange with the global scholarly community, -->
<!-- opens theory development to diverse perspectives, and enables (distributed and adversarial) collaboration. -->
